\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-26 paper.qmd
Added detailed description for simulated data example

In this section, we showcase the effectiveness of our methodology using simulated data. The dataset comprises five spherical Gaussian clusters in 4-$d$, with each cluster containing an equal number of points and consistent within variation.

In the 2D representations created by all NLDR techniques, as shown in @fig-nldervis5Gau, except for PHATE, there are five distinct clusters. In tSNE, these five clusters are closely located to each other (see @fig-nldervis5Gau (a)). However, in PHATE, two clusters are closely positioned, while the other three are more distant (see @fig-nldervis5Gau (c)). In PaCMAP, one cluster is at the center, and the remaining four are positioned in four different directions (see @fig-nldervis5Gau (e)). In TriMAP, two clusters are close, although not as close as in PHATE, and the other three are well-separated (see @fig-nldervis5Gau (d)). In UMAP, all clusters are arranged in a parallel manner, with three in one line and the other two in a separate line (see @fig-nldervis5Gau (b)).

Visualizing the models alongside the original high-D data provides insights into how different techniques capture the underlying clustering structure. The tSNE model exhibits five well-separated clusters, effectively preserving both local and global structures (see video link of @fig-modelfiveGau (a)). UMAP, on the other hand, also presents five distinct clusters but with a more flattened surface appearance (see video link of @fig-modelfiveGau (b)). The PHATE model shows five separated clusters resembling triangles or partial triangles but struggles to capture local structures (see video link of @fig-modelfiveGau (c)). TriMAP and PaCMAP both display five well-separated clusters with flat surfaces, each capturing within-cluster variation to varying extents. Despite the various 2D representations, all NLDR techniques preserve the global structure (see video link of @fig-modelfiveGau (d), (e) respectively). However, tSNE, effectively capture both local and global structures.

@@

```{r}
#| echo: false
#| fig-cap: "2D layouts from different NLDR techniques applied the same data: (a) tSNE (perplexity = 61), (b) UMAP (n_neighbors = 15), (c) PHATE (knn = 5), (d) TriMAP (n_inliers = 5, n_outliers = 4, n_random = 3), and (e) PaCMAP (n_neighbors = 10, init = random, MN_ratio = 0.9, FP_ratio = 2). Is there a best representation of the original data or are they all providing  equivalent information?"
#| label: fig-nldervis5Gau
#| out-width: 100%

@@

>>>>

In this section, the effectiveness of the algorithm is described using a simulated dataset. The dataset consists of five spherical Gaussian clusters in 4-$d$, with each cluster containing an equal number of points and the same within-cluster variation.

In the 2D layouts generated by various NLDR techniques, as shown in @fig-nldervis5Gau, five distinct clusters are observable, except for PHATE. In tSNE (see @fig-nldervis5Gau (a)), these clusters appear closely. UMAP arranges all clusters in a parallel manner, with three aligned in one line and the other two in a separate line (see @fig-nldervis5Gau (b)). In contrast, PHATE shows two closely positioned clusters and three more distant ones (see @fig-nldervis5Gau (c)). PaCMAP shows one central cluster and the remaining four spread out in different directions (see @fig-nldervis5Gau (e)). Finally, in TriMAP, two clusters are close, though not as tightly as PHATE, while the other three are well-separated (see @fig-nldervis5Gau (d)).

@@

```{r}
#| echo: false
#| fig-cap: "2D layouts from different NLDR techniques applied the same data: (a) tSNE (perplexity = 61), (b) UMAP (n_neighbors = 15), (c) PHATE (knn = 5), (d) TriMAP (n_inliers = 5, n_outliers = 4, n_random = 3), and (e) PaCMAP (n_neighbors = 10, init = random, MN_ratio = 0.9, FP_ratio = 2). Is there a best representation of the original data or are they all providing  equivalent information?"
#| label: fig-nldervis5Gau
#| fig-pos: H
#| out-width: 100%

@@

To investigate which is the best representation to visualize the original data or all NLDR methods provide equivalent information, we visualize the model-in-data space. Models from all NLDR methods show five well-separated clusters (see @fig-gau1_sc, @fig-gau2_sc, @fig-gau3_sc, @fig-gau4_sc, and @fig-gau5_sc). This suggests that for the five Gaussian cluster dataset, all NLDR methods effectively preserve the global structure. tSNE, UMAP, and PHATE display clusters with varying densities, indicating their ability to capture within-cluster variation (see @fig-gau1_sc, @fig-gau2_sc, and @fig-gau3_sc). On the other hand, both TriMAP and PaCMAP show clusters with flat surfaces, suggesting a failure to capture within-cluster variation (see @fig-gau4_sc and @fig-gau5_sc). Therefore, TriMAP and PaCMAP do not capture the local structure as effectively as other methods.

### 2024-03-26 paper.tex
Added detailed description for simulated data example

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

In this section, we showcase the effectiveness of our methodology using
simulated data. The dataset comprises five spherical Gaussian clusters
in 4-\(d\), with each cluster containing an equal number of points and
consistent within variation.

In the 2D representations created by all NLDR techniques, as shown in
Figure~\ref{fig-nldervis5Gau}, except for PHATE, there are five distinct
clusters. In tSNE, these five clusters are closely located to each other
(see Figure~\ref{fig-nldervis5Gau} (a)). However, in PHATE, two clusters
are closely positioned, while the other three are more distant (see
Figure~\ref{fig-nldervis5Gau} (c)). In PaCMAP, one cluster is at the
center, and the remaining four are positioned in four different
directions (see Figure~\ref{fig-nldervis5Gau} (e)). In TriMAP, two
clusters are close, although not as close as in PHATE, and the other
three are well-separated (see Figure~\ref{fig-nldervis5Gau} (d)). In
UMAP, all clusters are arranged in a parallel manner, with three in one
line and the other two in a separate line (see
Figure~\ref{fig-nldervis5Gau} (b)).

Visualizing the models alongside the original high-D data provides
insights into how different techniques capture the underlying clustering
structure. The tSNE model exhibits five well-separated clusters,
effectively preserving both local and global structures (see video link
of Figure~\ref{fig-modelfiveGau} (a)). UMAP, on the other hand, also
presents five distinct clusters but with a more flattened surface
appearance (see video link of Figure~\ref{fig-modelfiveGau} (b)). The
PHATE model shows five separated clusters resembling triangles or
partial triangles but struggles to capture local structures (see video
link of Figure~\ref{fig-modelfiveGau} (c)). TriMAP and PaCMAP both
display five well-separated clusters with flat surfaces, each capturing
within-cluster variation to varying extents. Despite the various 2D
representations, all NLDR techniques preserve the global structure (see
video link of Figure~\ref{fig-modelfiveGau} (d), (e) respectively).
However, tSNE, effectively capture both local and global structures.

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, frame hidden, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

In this section, the effectiveness of the algorithm is described using a
simulated dataset. The dataset consists of five spherical Gaussian
clusters in 4-\(d\), with each cluster containing an equal number of
points and the same within-cluster variation.

In the 2D layouts generated by various NLDR techniques, as shown in
Figure~\ref{fig-nldervis5Gau}, five distinct clusters are observable,
except for PHATE. In tSNE (see Figure~\ref{fig-nldervis5Gau} (a)), these
clusters appear closely. UMAP arranges all clusters in a parallel
manner, with three aligned in one line and the other two in a separate
line (see Figure~\ref{fig-nldervis5Gau} (b)). In contrast, PHATE shows
two closely positioned clusters and three more distant ones (see
Figure~\ref{fig-nldervis5Gau} (c)). PaCMAP shows one central cluster and
the remaining four spread out in different directions (see
Figure~\ref{fig-nldervis5Gau} (e)). Finally, in TriMAP, two clusters are
close, though not as tightly as PHATE, while the other three are
well-separated (see Figure~\ref{fig-nldervis5Gau} (d)).

@@

To investigate which is the best representation to visualize the
original data or all NLDR methods provide equivalent information, we
visualize the model-in-data space. Models from all NLDR methods show
five well-separated clusters (see Figure~\ref{fig-gau1_sc},
Figure~\ref{fig-gau2_sc}, Figure~\ref{fig-gau3_sc},
Figure~\ref{fig-gau4_sc}, and Figure~\ref{fig-gau5_sc}). This suggests
that for the five Gaussian cluster dataset, all NLDR methods effectively
preserve the global structure. tSNE, UMAP, and PHATE display clusters
with varying densities, indicating their ability to capture
within-cluster variation (see Figure~\ref{fig-gau1_sc},
Figure~\ref{fig-gau2_sc}, and Figure~\ref{fig-gau3_sc}). On the other
hand, both TriMAP and PaCMAP show clusters with flat surfaces,
suggesting a failure to capture within-cluster variation (see
Figure~\ref{fig-gau4_sc} and Figure~\ref{fig-gau5_sc}). Therefore,
TriMAP and PaCMAP do not capture the local structure as effectively as
other methods.
